---
title: "Project"
output:
  pdf_document: default
  html_document: default
date: "2026-04-09"
---

# How are housing property characteristics and location associated with housing prices?

We study how housing characteristics and location affect housing prices using Bayesian regression models. The response variable is log-transformed to reduce skewness and stabilize variance.

## Data cleaning
```{r}
library(ggplot2)
library(dplyr)

data <- read.csv("/Users/heleyi/Downloads/Bengaluru_House_Price_Data(main).csv")
head(data)

# 1. Remove column with too many missing values
data$society <- NULL

# 2. Extract numeric bedrooms from size
data$bedrooms <- as.numeric(gsub("[^0-9]", "", data$size))

# 3. Clean total_sqft
convert_sqft <- function(x) {
  if (grepl("-", x)) {
    nums <- as.numeric(unlist(strsplit(x, "-")))
    return(mean(nums))
  } else if (grepl("^[0-9.]+$", x)) {
    return(as.numeric(x))
  } else {
    return(NA)
  }
}

data$sqft <- sapply(data$total_sqft, convert_sqft)

# 4. Transform balcony 
data$balcony <- factor(data$balcony, levels = c(0,1,2,3))

# 5. Remove rows with missing key variables
data <- data[!is.na(data$sqft), ]
data <- data[!is.na(data$bath), ]
data <- data[!is.na(data$balcony), ]
data <- data[!is.na(data$bedrooms), ]
data <- data[!is.na(data$location), ]
data <- data[data$sqft > 0, ]

# 6. Frequency-based grouping for location
location_counts <- data %>%
  group_by(location) %>%
  tally(sort = TRUE)

top_18_names <- head(location_counts$location, 18)

data <- data %>%
  mutate(location_grouped = case_when(
    location %in% top_18_names ~ as.character(location),
    location_counts$n[match(location, location_counts$location)] >= 30 ~ "Other_Established",
    TRUE ~ "Other_Sparse"
  ))

# Convert to factor
all_levels <- c(as.character(top_18_names), "Other_Established", "Other_Sparse")
data$location_grouped <- factor(data$location_grouped, levels = all_levels)

# 7. Log-transform variables
data$log_price <- log(data$price)
data$log_sqft  <- log(data$sqft)  


# 8. Keep relevant variables
clean_data <- data[, c("log_price",
                       "log_sqft",
                       "balcony",
                       "location_grouped")]

# 9. Rename columns
colnames(clean_data) <- c("log_price",
                          "log_sqft",
                          "balcony",
                          "location")

# 10. Save cleaned dataset
write.csv(clean_data, "clean_housing.csv", row.names = FALSE)

# 11. Check data
head(clean_data)
table(clean_data$location)

```





# Model 1: Bayesian Linear Regression

## Data transformation

Both housing prices and square footage are typically right-skewed, so we apply log transformations to both variables:

\[
y_i = \log(\text{price}_i)
\]
\[
x_{i1} = \log(\text{total\_sqft}_i)
\]

- Balcony is treated as a categorical variable with four levels: \(\text{balcony}_i \in \{0,1,2,3\}\)
- We construct dummy variables with balcony = 0 as the reference category:

\[
D_{ik} = 
\begin{cases} 
1 & \text{if balcony}_i = k \\
0 & \text{otherwise}
\end{cases}
\]
for \(k \in \{1, 2, 3\}\).

We assume:
\[
y_i \sim \mathcal{N}(\mu_i, \sigma^2)
\]

Linear model

\[
\mu_i = \beta_0 + \beta_1 \log(\text{total\_sqft}_i) + \beta_2 D_{i1} + \beta_3 D_{i2} + \beta_4 D_{i3}
\]

## Priors

### Exploratory Data Analysis (EDA)
1. log(sqrt(sqft)) vs log(price)
```{r}

ggplot(data, aes(x = log_sqft, y = log_price)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "log(sqft) vs log(price)")
```

Interpretation: 
The slope is clearly positive
This indicates that as size increases, price increases monotonically
The relationship is smooth and strictly increasing

From exploratory data analysis (EDA), we observe a strong positive linear relationship between these log-transformed variables.

2. Balcony vs log(price)
```{r}
ggplot(data, aes(x = balcony, y = log_price)) +
  geom_boxplot() +
  labs(title = "Balcony vs log(price)")
```
Boxplots reveal that the effect of balconies is not strictly increasing (e.g., the median price for one balcony is slightly lower than for zero balconies). A Normal prior centered at zero allows the model to capture both positive and negative marginal effects.

Therefore, we set the prior distribution as:

\[
\beta_0 \sim \mathcal{N}(0, 10^2)
\]

\[
\beta_1 \sim \text{Gamma}(2, 1)
\]

\[
\beta_2, \beta_3, \beta_4 \sim \mathcal{N}(0, 2^2)
\]

\[
\sigma \sim \text{Exponential}(2)
\]

Interpretation

- \(\beta_0\): Baseline log price for properties with balcony = 0.
- \(\beta_1\): Strictly positive elasticity-like effect of square footage on price.
- \(\beta_2, \beta_3, \beta_4\): Discrete balcony effects relative to the baseline. This specification allows for non-linear trends where more balconies do not necessarily lead to higher prices.


# Model 2: Hierarchical Bayesian Model (Location Effects)

## Motivation

Housing prices vary significantly across locations. To capture spatial heterogeneity while avoiding overfitting, we introduce location-specific intercepts with partial pooling.

## Location structure

Given the long-tail distribution of housing locations (over 2,000 unique entries), we categorize them into 20 groups ($j[i] \in \{1, \dots, 20\}$) to balance spatial granularity with statistical reliability:

* Top 18 Locations: The 18 most frequent neighborhoods are modeled individually to capture their specific market premiums.
* Other (Established): Locations with 30 or more observations that did not make the top 18. This group represents relatively stable but less frequent areas.
* Other (Sparse): Locations with fewer than 30 observations. 

This group helps isolate high-variance noise and uncleaned data outliers, preventing them from biasing the core location estimates.

```{r}
ggplot(data, aes(x = reorder(location_grouped, log_price, FUN = median), y = log_price)) +
  geom_boxplot(aes(fill = location_grouped), outlier.alpha = 0.2) +
  coord_flip() +
  labs(title = "Log Price Distribution: Top 18 vs Refined Other Tiers",
       x = "Location Group",
       y = "Log Price") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Model specification

\[
y_i \sim \mathcal{N}(\mu_i, \sigma^2)
\]

\[
\mu_i = \alpha_{j[i]} + \beta_1 \log(\text{total\_sqft}_i) + \beta_2 D_{i1} + \beta_3 D_{i2} + \beta_4 D_{i3}
\]

### Hierarchical structure

\[
\alpha_j \sim \mathcal{N}(\mu_\alpha, \tau^2)
\]
\[
\mu_\alpha \sim \mathcal{N}(0, 10^2)
\]
\[
\tau \sim \text{Exponential}(2)
\]

### Coefficient priors

We maintain the specialized priors for the structural characteristics to ensure consistency between models:

\[
\beta_1 \sim \text{Gamma}(2, 1)
\]
\[
\beta_2, \beta_3, \beta_4 \sim \mathcal{N}(0, 2^2)
\]
\[
\sigma \sim \text{Exponential}(2)
\]

## Interpretation

- \(\alpha_j\): Baseline price level for each location group. The hierarchical structure allows for "information borrowing," where sparse locations are pulled toward the global mean $\mu_\alpha$.
- \(\tau\): Degree of variation across different housing markets.
- \(\beta_2, \beta_3, \beta_4\): Discrete balcony effects adjusted for location-specific baselines.

